Nested Stochastic Block Models Applied to the Analysis of Single Cell Data

### 2020-11-16 manuscript.tex
initial import of old files

% Template for PLoS
% Version 3.5 March 2018
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8x]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
%\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
%\setlength{\headheight}{27.023pt}
%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\today}

%% Include all macros below

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

%% END MACROS SECTION

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Nested stochastic block models applied to the analysis of single cell data} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Leonardo Morelli\textsuperscript{1,2},
Valentina Giansanti\textsuperscript{1, 3},
Davide Cittaro\textsuperscript{1*},
\\
\bigskip
\textbf{1} Center for Omics Sciences, IRCCS San Raffaele Institute, Milan, Italy
\\
\textbf{2} Universit\`{a} Vita-Salute San Raffaele, Milan, Italy
\\
\textbf{3} Department of Informatics, Systems and Communication, University of Milano-Bicocca, Milan, Italy
\\
\bigskip

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
% 
% Remove or comment out the author notes below if they aren't used.
%
% Primary Equal Contribution Note
%\Yinyang These authors contributed equally to this work.

% Additional Equal Contribution Note
% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
%\ddag These authors also contributed equally to this work.

% Current address notes
%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
% \textcurrency b Insert second current address 
% \textcurrency c Insert third current address

% Deceased author note
%\dag Deceased

% Group/Consortium Author Note
%\textpilcrow Membership list can be found in the Acknowledgments section.

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* cittaro.davide@hsr.it

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Single cell profiling has been proven to be a powerful tool in molecular biology to understand the complex behaviours of heterogeneous system. While properties of single cells is the primary endpoint of such analysis, these are typically clustered to underpin the common determinants that can be used to describe functional properties of the cell mixture under investigation. Several approaches have been proposed to identify cell clusters; while this is matter of active research, one popular approach is based on community detection in neighbourhood graphs by optimisation of modularity. In this paper we propose an alternative solution to this problem, based on nested Stochastic Block Models; we show a threefold advantage of our approach as it is able to correctly identify cell groups, it returns a meaningful hierarchical structure and, lastly, it provides a statistical measure of association between cells and the assigned clusters.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
Identification of cell types is a key step of many single cell experiments. Many of the approaches to achieve this goal converge on the analysis of the graph representing cell-wise similarity by optimisation of modularity. While fast, this method does not provide a robust estimation of the number of groups, and it leaves large room of arbitrariness in the choice of appropriate resolution. To overcome these limitations, we propose a strategy based on nested stochastic block models. We show our method is accurate and provides a rich description of cell groups and their relationships.

\linenumbers

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}
Transcriptome analysis at single cell level by RNA sequencing (scRNA-seq) is a technology growing in popularity and applications \cite{svensson_2018}. It has been applied to study the biology of complex tissues \cite{guo_2018, ventotormo_2018}, tumor dynamics \cite{rozenblattrosen_2020, tirosh_2016, patel_2014, neftel_2019}, development \cite{rosenberg_2018, wagner_2018} and to describe whole organisms \cite{plass_2018, regev_2017}.

@@

%\begin{MPEquation}[!ht]
\begin{equation}
P(\boldsymbol b | \boldsymbol A) = \frac{P(\boldsymbol A|\boldsymbol\theta, \boldsymbol b)P(\boldsymbol\theta, \boldsymbol b)}{P(\boldsymbol A)}
\end{equation}
\label{MPEquationElement:820CB015-E51E-4C5A-BD9C-70E667B8F73B}
%\end{MPEquation}
and inference is performed by maximising the posterior probability. The numerator in this equation can be rewritten exponentiating the description length

%\begin{MPEquation}[!ht]
\begin{equation}
\Sigma = -\ln P(\boldsymbol A|\boldsymbol\theta, \boldsymbol b) - \ln P(\boldsymbol\theta, \boldsymbol b)
\end{equation}
\label{MPEquationElement:676396C0-2D6F-4084-89B1-951DE3032087}
%\end{MPEquation}
so that inference is performed by minimizing the information required to describe the data (Occam's razor); \emph{graph-tool} is able to efficiently do this by a Markov Chain Monte Carlo approach \cite{peixoto_2014}. SBM itself may fail to identify small groups in large graphs, hence hierarchical formulation has been proposed \cite{peixoto_2014_h}. Under this model, communities are agglomerated at a higher level in a block multigraph, also modelled using SBM. This process is repeated recursively until a graph with a single block is reached, creating a Nested Stochastic Block Model (nSBM).

@@

\section*{Materials and methods}
%\subsection*{Analysis of Random data}
%Data were retrieved in \emph{scanpy} environment using \emph{sc.datasets.pbmc3k\_processed()} function. The random \emph{k}NN graph was obtained shuffling the node labels of each edge. nSBM was performed with 3 initialisations using merge-split MCMC approach. UMAP embedding was recomputed after randomisation using the shuffled graph.

@@

Data and metadata for five cell mixture profiled by Chromium 10x were downloaded from the sc-mixology repository (\href{https://github.com/LuyiTian/sc_mixology}{https:/\slash github.com\slash LuyiTian\slash sc\_mixology}). Data were analysed using scanpy v1.4.6 \cite{wolf_2018}. Cells with less than 200 genes were excluded, as genes detected in less than 3 cells. Cells with less than 5\% of mitochondrial genes were retained for subsequent analysis. Data were normalised and log-transformed; number of genes and percentage of mitochondrial genes were regressed out. nSBM was initialised three times. Analysis was performed at level 2 of the nSBM hierarchy.
Random Poisson noise was generated at given $\lambda$ (range: 0 - 1000) and added to the initial count matrix. Analysis was performed at level 2 of the nSBM hierarchy.

@@

% Results and Discussion can be combined.
\subsection*{Overview of \emph{schist}}

%\subsection*{nSBM does not return artificial cell groupings}
%One of the most relevant difference between nSBM and other methods to cluster single cells is that it relies on robust statistical modelling. In this sense, the number of groups identified strictly mirrors the amount of information contained in the data. An important consequence is that absence of information can be properly handled. To show this property we performed a simple experiment on a randomised \emph{k}NN graph. To this end we collected data for 3k PBMC (available as preprocessed data in \emph{scanpy}) and shuffled the edges of the prebuilt \emph{k}NN graph, this to keep the general graph properties unchanged. We tested that the degree distribution does not change after randomisation (Kolmogorov-Smirnov $D=0.0431$, $p=0.993$). We found that the default strategy, based on maximisation of modularity, identifies 24 cell groups, whereas nSBM does not identify any cell group.

%This experiment is a deliberate extreme case. The quality of grouping proposed by a standard approach can be disputed in many ways, and the UMAP embedding indeed reflects the absence of any information. Nevertheless, real-world data may include an unknown amount of random noise. Hence, It is important to identify cell groups that are not artefacts arising from processing and that do not reflect the information contained in the dataset.

%\begin{figure}[!h]
%\centering
%\includegraphics[keepaspectratio,width=\textwidth,height=0.4\textheight]{FIgure_Random.png}
%\caption[]{nSBM results for randomised data. (A) UMAP embedding of ~3k PBMC after standard processing with Leiden approach (left) or nSBM (right). Cell grouping is consistent for both the approaches ($ARI=0.804$). (B) UMAP embedding of the same data after randomisation of \emph{k}NN graph edges. In this case nSBM does not return any cell grouping, while optimisation of modularity finds up to 24 different cell groups}\label{FigureRandom}
%\end{figure}

\subsection*{nSBM correctly identifies cell populations}
To benchmark \emph{schist}, we tested our approach on scRNA-seq mixology data \cite{tian_2019}, in particular on a mixture of 5 cell lines profiled with Chromium 10x platform. At a first evaluation of the UMAP embedding, all lines appear well separated. Only the lung cancer line H1975 shows a certain degree of heterogeneity with some cells being embedded in other cell groups (Fig. \ref{Figure1}A). Inference on the neighbourhood graph is influenced by the graph structure itself, therefore we built multiple graphs changing the number of principal components used in PCA reduction and the number of neighbours in the \emph{k}NN graph. We then calculated the Adjusted Rand Index (\emph{ARI}) between the cell line assignments (ground truth) and the cell groups identified by nSBM at each level. We found a peak of $ARI=0.977$ with 30 principal components (PC) and 30 neighbours. In general, higher number of components and neighbours has a positive impact on the performance (Fig. \ref{Figure1}B). Conversely, if few PCs (10) or neighbors (5) are used, performances degrade, with a minimum $ARI=0.669$ at 20 PCs and 5 neighbours. If fewer PCs are used, a smaller fraction of the total variance, hence less information, is used to build the \emph{k}NN graph; if fewer neighbours are chosen, the graph is sparser and the model is fit from less edges (\nameref{S1_Fig}). Running MCMC algorithm recovers the performances of the majority of the configurations (\nameref{S2_Fig}).

It is worth noting that since nSBM reflects the amount of information contained in the data, it does not return clusters when data are noisy. To demonstrate this property, we added increasing levels of random Poisson noise to the count matrix (Fig.\ref{Figure_Random}). The number of groups reported by nSBM decreases with increasing amount of noise and it is 1 (\emph{i.e.} no clusters) at $\lambda=200$, whereas a standard approach overfits data.

These observations show that nSBM is suitable for accurate identification of cell groups, without the need of an arbitrary threshold on the resolution parameter and with the possibility to identify rare cell types in larger populations. Moreover, nSBM avoids excessive clustering of data in presence of noisy measurements.

\begin{figure}[!h]
\centering
%\includegraphics[keepaspectratio,width=\textwidth,height=0.6\textheight]{Figure_1.png}
\caption[]{\emph{schist} applied to scRNA-seq mixology data. (A) UMAP embedding of 10x Chromium data, cells are colored according to the given cell line in the original paper. A small number of H1975 cells are found in HCC827 and H838 clusters. (B) Heatmap showing the maximal Adjusted Rand Index for different \emph{k}NN graphs. We tested the impact of varying the number of Principal Components and the number of neighbors used in \emph{sc.pp.neighbors()} function in \emph{scanpy}. Adjusted Rand Index between the actual cell lines and the identified groups is shown. Darker blue indicates higher concordance between the model and the ground truth. (C) Alluvial plots showing the hierarchy of cell groups as identified by \emph{schist} (above) or by Leiden method at different resolution thresholds (below). The bars on the right indicate the cell identity; two marks in the \emph{schist} plot indicate two groups of cells discussed in the main text}\label{Figure1}
\end{figure}

\begin{figure}[!h]
\centering
%\includegraphics[keepaspectratio,width=\textwidth,height=0.6\textheight]{Tian_Random.png}
\caption[]{Identification of cell groups at different level of random noise. When random Poisson noise is added to the gene count martix, at increasing values of $\lambda$, a strategy based on optimization of modularity does return cell groups, whereas nSBM does not cluster cells when the noise exceeds a certain threshold (solid lines). The behaviour of the former approach follows the trend of the modularity (dashed lines), which is sustained at high levels independently from the level of noise.}\label{Figure_Random}
\end{figure}

\subsection*{Model hierarchy contains biological information}
The hierarchical model of cell groups implies that a relationship exists between groups. We next wanted to explore if the hierarchy proposed by the nSBM had a biological interpretation. To this end, we analysed data for hematopoietic differentiation \cite{paul_2015}, previously used to benchmark the consistency of cell grouping with differentiation trajectories by graph abstraction \cite{wolf_2019}. Standard processing of those data reveals three major branchings (Erythroids, Neutrophils and Monocytes) stemming from the progenitor cells (\nameref{S3_Fig}A). After applying nSBM, we identify 27 groups at level 3 of the hierarchy (\nameref{S3_Fig}B), compared to the 24 using Leiden method at default resolution (\nameref{S4_Fig}). We found that the hierarchy proposed by our model is consistent with the developmental model (Fig. \ref{Figure2}). Of note, we found that clustering with Leiden method produces cell groups that are mixed and split at different resolutions (0.1 - 1), in a non hierarchical manner (\nameref{S4_Fig}); we spotted several occurrences of such phenomenon, \emph{e.g.} group 9 at resolution $r=0.4$ splits into groups 0 and 6 at $r=0.3$ or group 3 at $r=0.6$ splits in groups 4, 8 and 12 at $r=0.5$.

\begin{figure}[!h]
\centering
%\includegraphics[keepaspectratio,width=\textwidth,height=0.55\textheight]{Figure_2.png}
\caption[]{Analysis of hematopoietic differentiation. Each panel presents a low dimensional embedding of single cells next to a radial tree representation of the nSBM hierarchy. Cells are colored according to groupings at level 5 of the hierarchy, group 0 marks the progenitor population (A). In subsequent panels, cells are colored using a signature of erythroid lineage (B), monocytes (C) or neutrophils (D).}\label{Figure2}
\end{figure}

Cluster consistency is not a measure of the data quality, in fact we identify low consistency groups in all datasets. High consistency, instead, appears to be linked to the biological purity of the cells and it is inverse to the diversity index, estimated using cell annotation from the original paper. Consequently, filtering low consistency groups increases concordance with biological groups, at the cost of a reduced number of cells ( \nameref{S5_Fig}).

\begin{figure}[!h]
\centering
%\includegraphics[keepaspectratio,width=\textwidth,height=0.6\textheight]{Figure_3.png}
\caption[]{Analysis of cell cluster consistency. Every panel reports a UMAP embedding of a PBMC + HEK293 cells profiled on different platform. Cells are annotated by cell type and by consistency value, which is assigned to cell clusters at nSBM level 1. The charts next to UMAPs show the correlation between consistency and diversity index for each cell cluster. Technologies showed here are (A) Chromium 10x v3, (B) Quartz-seq 2, (C) MARS-seq and (D) iCELL8.}\label{Figure3}
\end{figure}

\section*{Conclusion}

\section*{Supporting information}

% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
\paragraph*{S1 Fig.}
\label{S1_Fig}
{\bf Degree distribution of \emph{k}NN graphs.} Degree distribution of multiple \emph{k}NN graphs derived from scRNA-seq mixology datasets using variable number of Principal Components or number of neighbors. Each histogram shows the number of nodes (on y axis) within a specific degree bin (on x axis). Both the parameters influence the sparseness of the graph.

\paragraph*{S2 Fig.}
\label{S2_Fig}
{\bf nSBM performance after MCMC.}  Adjusted Rand Index for different \emph{k}NN graphs after MCMC run. Maximal ARI over all hierarchy level is shown. Darker color indicates higher concordance with the ground truth.

>>>>

\documentclass[10pt]{article}
\usepackage[top=20mm, bottom=20mm, left=20mm, right=20mm]{geometry}
\geometry{a4paper}
\geometry{portrait}
\usepackage[utf8]{inputenc}	% Comment out for xelatex; must use with pdflatex
\usepackage[T1]{fontenc}
\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{fontspec}
\usepackage{epstopdf}
\usepackage{float}
\usepackage[hang]{footmisc}
\usepackage{endnotes}
\usepackage{listingsutf8}
%\restylefloat{table}
\usepackage{xcolor}
\usepackage{colortbl}
%\usepackage{booktabs}
\usepackage{tabu}
\usepackage{caption}
\usepackage{textcomp}
\usepackage{textgreek}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage[normalem]{ulem}
\usepackage{authblk}
\usepackage[numbers]{natbib}
\bibliographystyle{plos2015}

\makeatletter
\def\blx@maxline{77}
\makeatother

\captionsetup[figure]{labelfont=sc}
\captionsetup[table]{labelfont=sc}
\captionsetup[tabu]{labelfont=sc}

\DeclareCaptionType{MPEquation}[][List of equations]
\captionsetup[MPEquation]{labelformat=empty}

\DeclareCaptionType{MPListing}[][List of listings]
\captionsetup[MPListing]{labelformat=empty}

\usepackage{fancyhdr}
\setlength{\headheight}{16pt}
\pagestyle{fancyplain}
\fancyhf{}
%\fancypagestyle{plain}{}
\cfoot[\thepage]{\thepage}

\usepackage{stringenc}
\usepackage{pdfescape}

% Helper for debugging pdflatex errors induced by invalid UTF8 characters:
% output the actual character code causing the issue.

\makeatletter
\renewcommand*{\UTFviii@defined}[1]{%
\ifx#1\relax
\begingroup
% Remove prefix "\u8:"
\def\x##1:{}%
% Extract Unicode char from command name
% (utf8.def does not support surrogates)
\edef\x{\expandafter\x\string#1}%
\StringEncodingConvert\x\x{utf8}{utf16be}% convert to UTF-16BE
% Hexadecimal representation
\EdefEscapeHex\x\x
% Enhanced error message
\PackageError{inputenc}{Unicode\space char\space \string#1\space
(U+\x)\MessageBreak
not\space set\space up\space
for\space use\space with\space LaTeX}\@eha
\endgroup
\else\expandafter
#1%
\fi
}
\makeatother

% Make multi-paragraph footnotes align nicely on the left
\setlength\footnotemargin{10pt}

% Do not display an automatic Notes heading for endnotes, as they are contained
% by a Manuscripts section with a heading that has possibly been edited by the user
\def\enoteheading{}

% Apply footnote and endnote numbering schemes (decimal, lowercase roman etc.)
\renewcommand{\thefootnote}{\arabic{footnote}}
\renewcommand{\theendnote}{\arabic{endnote}}
% Adjust endnotes to use normal-sized numbering and text
\renewcommand{\enotesize}{\normalsize}
\renewcommand\enoteformat{%
  \raggedright
  \leftskip=1.8em
  \makebox[0pt][r]{\theenmark. \rule{0pt}{\dimexpr\ht\strutbox+\baselineskip}}%
}

\begin{document}
\title{Nested Stochastic Block Models Applied to the Analysis of Single Cell Data}
\author[1,2]{Leonardo Morelli}
\author[1,3]{Valentina Giansanti}
\author[1]{Davide Cittaro}
\affil[1]{Center for Omics Sciences, IRCCS San Raffaele Institute, Milan, Italy}
\affil[2]{Universit\`{a} Vita-Salute San Raffaele, Milan, Italy} 
\affil[3]{Department of Informatics, Systems and Communication, University of Milano-Bicocca, Milan, Italy}
\maketitle

\section*{Abstract}

Single cell profiling has been proven to be a powerful tool in molecular biology to understand the complex behaviours of heterogeneous system. While properties of single cells is the primary endpoint of such analysis, these are typically clustered to underpin the common determinants that can be used to describe functional properties of the cell mixture under investigation. Several approaches have been proposed to identify cell clusters; while this is matter of active research, one popular approach is based on community detection in neighbourhood graphs by optimisation of modularity. In this paper we propose an alternative solution to this problem, based on nested Stochastic Block Models; we show a threefold advantage of our approach as it is able to correctly identify cell groups, it returns a meaningful hierarchical structure and, lastly, it provides a statistical measure of association between cells and the assigned clusters.

\section*{Background}

Transcriptome analysis at single cell level by RNA sequencing (scRNA-seq) is a technology growing in popularity and applications \cite{svensson_2018}. It has been applied to study the biology of complex tissues \cite{guo_2018, ventotormo_2018}, tumor dynamics \cite{rozenblattrosen_2020, tirosh_2016, patel_2014, neftel_2019}, development \cite{rosenberg_2018, wagner_2018} and to describe whole organisms \cite{plass_2018, regev_2017}.

@@

\begin{MPEquation}[!ht]
\begin{equation}
P(\boldsymbol b | \boldsymbol A) = \frac{P(\boldsymbol A|\boldsymbol\theta, \boldsymbol b)P(\boldsymbol\theta, \boldsymbol b)}{P(\boldsymbol A)}
\end{equation}
\label{MPEquationElement:820CB015-E51E-4C5A-BD9C-70E667B8F73B}
\end{MPEquation}
and inference is performed by maximising the posterior probability. The numerator in this equation can be rewritten exponentiating the description length

\begin{MPEquation}[!ht]
\begin{equation}
\Sigma = -\ln P(\boldsymbol A|\boldsymbol\theta, \boldsymbol b) - \ln P(\boldsymbol\theta, \boldsymbol b)
\end{equation}
\label{MPEquationElement:676396C0-2D6F-4084-89B1-951DE3032087}
\end{MPEquation}
so that inference is performed by minimizing the information required to describe the data (Occam's razor); \emph{graph-tool} is able to efficiently do this by a Markov Chain Monte Carlo approach \cite{peixoto_2014}. SBM itself may fail to identify small groups in large graphs, hence hierarchical formulation has been proposed \cite{peixoto_2014_h}. Under this model, communities are agglomerated at a higher level in a block multigraph, also modelled using SBM. This process is repeated recursively until a graph with a single block is reached, creating a Nested Stochastic Block Model (nSBM).

@@

\section*{Results}

\subsection*{Overview of \emph{schist}}

\subsection*{nSBM does not return artificial cell groupings}
One of the most relevant difference between nSBM and other methods to cluster single cells is that it relies on robust statistical modelling. In this sense, the number of groups identified strictly mirrors the amount of information contained in the data. An important consequence is that absence of information can be properly handled. To show this property we performed a simple experiment on a randomised \emph{k}NN graph. To this end we collected data for 3k PBMC (available as preprocessed data in \emph{scanpy}) and shuffled the edges of the prebuilt \emph{k}NN graph, this to keep the general graph properties unchanged. We tested that the degree distribution does not change after randomisation (Kolmogorov-Smirnov $D=0.0431$, $p=0.993$). We found that the default strategy, based on maximisation of modularity, identifies 24 cell groups, whereas nSBM does not identify any cell group.

\begin{figure}[H]
\centering
\includegraphics[keepaspectratio,width=\textwidth,height=0.4\textheight]{FIgure_Random.png}
\caption[]{nSBM results for randomised data. (A) UMAP embedding of ~3k PBMC after standard processing with Leiden approach (left) or nSBM (right). Cell grouping is consistent for both the approaches ($ARI=0.804$). (B) UMAP embedding of the same data after randomisation of \emph{k}NN graph edges. In this case nSBM does not return any cell grouping, while optimisation of modularity finds up to 24 different cell groups}\label{FigureRandom}
\end{figure}

This experiment is a deliberate extreme case. The quality of grouping proposed by a standard approach can be disputed in many ways, and the UMAP embedding indeed reflects the absence of any information. Nevertheless, real-world data may include an unknown amount of random noise. Hence, It is important to identify cell groups that are not artefacts arising from processing and that do not reflect the information contained in the dataset.

\subsection*{nSBM correctly identifies cell populations}
To benchmark \emph{schist}, we tested our approach on scRNA-seq mixology data \cite{tian_2019}, in particular on a mixture of 5 cell lines profiled with Chromium 10x platform. At a first evaluation of the UMAP embedding, all lines appear well separated. Only the lung cancer line H1975 shows a certain degree of heterogeneity with some cells being embedded in other cell groups (Fig. \ref{Figure1}A). Inference on the neighbourhood graph is influenced by the graph structure itself, therefore we built multiple graphs changing the number of principal components used in PCA reduction and the number of neighbours in the \emph{k}NN graph. We then calculated the Adjusted Rand Index (\emph{ARI}) between the cell line assignments (ground truth) and the cell groups identified by nSBM at each level. We found a peak of $ARI=0.977$ with 30 principal components (PC) and 30 neighbours. In general, higher number of components and neighbours has a positive impact on the performance (Fig. \ref{Figure1}B). Conversely, if few PCs (10) or neighbors (5) are used, performances degrade, with a minimum $ARI=0.669$ at 20 PCs and 5 neighbours. If fewer PCs are used, a smaller fraction of the total variance, hence less information, is used to build the \emph{k}NN graph; if fewer neighbours are chosen, the graph is sparser and the model is fit from less edges (Fig. S1). Running MCMC algorithm recovers the performances of the majority of the configurations (Fig. S2).

\begin{figure}[H]
\centering
\includegraphics[keepaspectratio,width=\textwidth,height=0.6\textheight]{Figure_1.png}
\caption[]{\emph{schist} applied to scRNA-seq mixology data. (A) UMAP embedding of 10x Chromium data, cells are colored according to the given cell line in the original paper. A small number of H1975 cells are found in HCC827 and H838 clusters. (B) Heatmap showing the maximal Adjusted Rand Index for different \emph{k}NN graphs. We tested the impact of varying the number of Principal Components and the number of neighbors used in \emph{sc.pp.neighbors()} function in \emph{scanpy}. Adjusted Rand Index between the actual cell lines and the identified groups is shown. Darker blue indicates higher concordance between the model and the ground truth. (C) Alluvial plots showing the hierarchy of cell groups as identified by \emph{schist} (above) or by Leiden method at different resolution thresholds (below). The bars on the right indicate the cell identity; two marks in the \emph{schist} plot indicate two groups of cells discussed in the main text}\label{Figure1}
\end{figure}

These observations show that nSBM is able to perform accurate identification of cell groups, without the need of an arbitrary threshold on the resolution parameter. These data also hint at the possibility to identify rare cell types in larger populations.

\begin{figure}[H]
\centering
\includegraphics[keepaspectratio,width=\textwidth,height=0.6\textheight]{Figure_3.png}
\caption[]{Analysis of cell cluster consistency. Every panel reports a UMAP embedding of a PBMC + HEK293 cells profiled on different platform. Cells are annotated by cell type and by consistency value, which is assigned to cell clusters at nSBM level 1. The charts next to UMAPs show the correlation between consistency and diversity index for each cell cluster. Technologies showed here are (A) Chromium 10x v3, (B) Quartz-seq 2, (C) MARS-seq and (D) iCELL8.}\label{Figure3}
\end{figure}

Cluster consistency is not a measure of the data quality, in fact we identify low consistency groups in all datasets. High consistency, instead, appears to be linked to the biological purity of the cells and it is inverse to the diversity index, estimated using cell annotation from the original paper. Consequently, filtering low consistency groups increases concordance with biological groups, at the cost of a reduced number of cells (Fig. S5).

\subsection*{Model hierarchy contains biological information}
The hierarchical model of cell groups implies that a relationship exists between groups. We next wanted to explore if the hierarchy proposed by the nSBM had a biological interpretation. To this end, we analysed data for hematopoietic differentiation \cite{paul_2015}, previously used to benchmark the consistency of cell grouping with differentiation trajectories by graph abstraction \cite{wolf_2019}. Standard processing of those data reveals three major branchings (Erythroids, Neutrophils and Monocytes) stemming from the progenitor cells (Fig. S3A). After applying nSBM, we identify 27 groups at level 3 of the hierarchy (Fig. S3B), compared to the 24 using Leiden method at default resolution (Fig. S4). We found that the hierarchy proposed by our model is consistent with the developmental model (Fig. \ref{Figure2}). Of note, we found that clustering with Leiden method produces cell groups that are mixed and split at different resolutions (0.1 - 1), in a non hierarchical manner (Fig. S4); we spotted several occurrences of such phenomenon, \emph{e.g.} group 9 at resolution $r=0.4$ splits into groups 0 and 6 at $r=0.3$ or group 3 at $r=0.6$ splits in groups 4, 8 and 12 at $r=0.5$.

\begin{figure}[H]
\centering
\includegraphics[keepaspectratio,width=\textwidth,height=0.55\textheight]{Figure_2.png}
\caption[]{Analysis of hematopoietic differentiation. Each panel presents a low dimensional embedding of single cells next to a radial tree representation of the nSBM hierarchy. Cells are colored according to groupings at level 5 of the hierarchy, group 0 marks the progenitor population (A). In subsequent panels, cells are colored using a signature of erythroid lineage (B), monocytes (C) or neutrophils (D).}\label{Figure2}
\end{figure}

\section*{Conclusions}

\section*{Materials and Methods}

\subsection*{Analysis of Random data}
Data were retrieved in \emph{scanpy} environment using \emph{sc.datasets.pbmc3k\_processed()} function. The random \emph{k}NN graph was obtained shuffling the node labels of each edge. nSBM was performed with 3 initialisations using merge-split MCMC approach. UMAP embedding was recomputed after randomisation using the shuffled graph.

@@

Data and metadata for five cell mixture profiled by Chromium 10x were downloaded from the sc-mixology repository (\href{https://github.com/LuyiTian/sc_mixology}{https:/\slash github.com\slash LuyiTian\slash sc\_mixology}). Data were analysed using scanpy v1.4.6 \cite{wolf_2018}. Cells with less than 200 genes were excluded, as genes detected in less than 3 cells. Cells with less than 5\% of mitochondrial genes were retained for subsequent analysis. Data were normalised and log-transformed; number of genes and percentage of mitochondrial genes were regressed out. nSBM was initialised three times

@@

\section*{Acknowledgements}

We would like to thank Tiago de Paula Peixoto (Central European University, ISI Foundation) and Giovanni Petri (ISI Foundation) for the discussions and the precious hints. We also would like to thank all people at COSR, in particular Giovanni Tonon and Paolo Provero.
This work has been supported by Accelerator Award: A26815 entitled:  "Single-cell cancer evolution in the clinic" funded through a partnership between Cancer Research UK and Fondazione AIRC

\bibliography{references}
